---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This is my report for the first home assignment in "Reproducible Research".

## Loading and preprocessing the data
In a first step, the zipped data is unzipped into a subdirectory called "data". Then, it is loaded into the workspace
and stored as a data.table.

```{r}
library(data.table)
unzip(zipfile = 'activity.zip', exdir = './data')
activity_data <- fread('./data/activity.csv', colClasses = c('integer', 'date', 'integer'))

```
The date column is converted into date format:
```{r}
activity_data$date <- as.Date(activity_data$date, format = "%Y-%m-%d")
head(activity_data)
```


## What is mean total number of steps taken per day?
To answer this question, the data has to be aggregated per day:

```{r}
agg_data_day <- activity_data[,list(steps_per_day = sum(steps, na.rm = TRUE)), by=date]
mean = mean(agg_data_day$steps_per_day, na.rm = TRUE)
median = median(agg_data_day$steps_per_day, na.rm = TRUE)
head(agg_data_day)
```

**Median** of the total number of steps is **`r median`** and **mean** is **`r mean`**. The following histogram shows the frequency of total steps per day over the considered period:

```{r hist_steps_day, fig.height = 4}
hist(agg_data_day$steps_per_day, breaks = 10, main = "Histogram of steps per day", 
     xlab = "Total number of steps per day")

```


## What is the average daily activity pattern?

Reaggregate the data to get the average number of steps per time interval:

```{r plot_steps_interval}
agg_data_timestep <- activity_data[,list(avg_steps_per_time = mean(steps, na.rm = TRUE)), by=interval]
plot(agg_data_timestep$interval, agg_data_timestep$avg_steps_per_time, type = "l",
     xlab = "Interval of day", ylab = "Mean number of steps", 
     main = "Average of steps per time interval")
agg_data_timestep <- agg_data_timestep[order(-avg_steps_per_time)]
max_interval <- agg_data_timestep[1]
```

The interval with the most steps on average is **interval `r max_interval$interval`** with
an **average number of steps of `r max_interval$avg_steps_per_time`.**


## Imputing missing values
First, calculate the number of rows with NAs:

```{r}
nas <- is.na.data.frame(activity_data)
print(summary(nas))
```

As indicated above, the only column with missing values is the "steps" column. As a strategy to impute missing
values, I therefore choose to replace the NAs by the corresponding average values of that time interval over 
all days.

```{r}
activity_data_cleaned <- merge(activity_data, agg_data_timestep, by="interval")
activity_data_cleaned <- activity_data_cleaned[,steps_cleaned:=ifelse(is.na(steps),avg_steps_per_time, steps)]
activity_data_cleaned <- activity_data_cleaned[,`:=`(steps=NULL, avg_steps_per_time=NULL)]
agg_data_day_cleaned <- activity_data_cleaned[,list(steps_per_day = sum(steps_cleaned)), by=date]
mean_cleaned = mean(agg_data_day_cleaned$steps_per_day)
median_cleaned = median(agg_data_day_cleaned$steps_per_day)
head(agg_data_day_cleaned)
```

The **median** of the cleaned steps per day is now **`r median_cleaned`** and the **mean** is **`r mean_cleaned`**. 
Apparently, the pre-processing by imputing missing values has a significant impact on the estimates of total daily
number of steps as this histogram shows in comparison to the one above:

```{r hist_steps_day_cleaned}
hist(agg_data_day_cleaned$steps_per_day, breaks = 10, main = "Histogram of steps per day",
     xlab = "Total number of steps per day")
```


## Are there differences in activity patterns between weekdays and weekends?

To answer this question, a new variable "day_type" is created
```{r}
activity_data_cleaned <- activity_data_cleaned[,day_type:=as.factor(ifelse(weekdays(date) %in% c("Samstag", "Sonntag"),"weekend","weekday"))]
head(activity_data_cleaned)
```

```{r plot_steps_interval_day_type}
library(ggplot2)
agg_data_timestep_daytype <- activity_data_cleaned[,list(avg_steps_per_time = mean(steps_cleaned, na.rm = TRUE)), by=list(interval, day_type)]
g <- ggplot(agg_data_timestep_daytype, aes(interval, avg_steps_per_time)) +
        geom_line(colour = "blue") + 
        facet_wrap(~day_type) + 
        labs(x = "Interval of day", y = "Mean number of steps")
g
```

Obviously, activity patterns differ depending on whether a weekday or weekend is considered. 
